---
title: "Investigation of exp distribution in R"
author: "Sukalpo Mitra"
date: "Sunday, May 24, 2015"
output: pdf_document
---

#Overview

In this project we will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. lambda is set to 0.2 for all of the simulations. We will investigate the distribution of averages of 40 exponentials. Note that we will need to do a thousand simulations.

#Simulation

```{r}
set.seed(820)
lambda = 0.2
x <- apply(matrix(rexp(40000,lambda),1000),1,mean)
```

The above code stores 1000 averages, each of a sample of 40 random exponential with lambda as the rate parameter in a variable called x.

#Sample Mean versus Theoretical Mean:

Sample mean is as follows 

```{r}
round(mean(x))
```

The theoritical mean is

```{r}
1/lambda
```

The following histogram shows the theoritical and analytical mean

```{r}
hist(x, xlab = "mean", main = "Exponential Distribution")
abline(v = round(mean(x)), col = "red")
abline(v = 1/lambda, col = "blue")
```

#Sample Variance versus Theoretical Variance:

Sample variance is as follows

```{r}
round(var(x),1)
```

Theoritical variance is as follows

```{r}
round((1/lambda/sqrt(40))^2,1)
```

The following histogram shows the theoritical and analytical variance

```{r}
hist(x, xlab = "variance", main = "Exponential Distribution")
abline(v = round(var(x),2) + mean(x), col = "red")
abline(v = mean(x) - round(var(x),2), col = "red")
abline(v = round((1/lambda/sqrt(40))^2,2) + mean(x), col = "blue")
abline(v = mean(x) - round((1/lambda/sqrt(40))^2,2), col = "blue")
legend("topright", c("Theoritical Variance", "Analytical Variance"), col=c("blue", "red"), lwd=10)
```

The variance when rounded to 2 decimal places are
```{r}
round(var(x),2)
round((1/lambda/sqrt(40))^2,2)
```

The difference is within 6% which is very less and as well when rounded to 1 decimal place it's almost same.

#Distribution

```{r}
xfit <- seq(min(x), max(x), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(40)))
hist(x,breaks=40,prob=T,col="red",xlab = "means",main="Distribution",ylab="density")
lines(xfit, yfit, pch=22, col="black", lty=5)
```